---
title: "In-class Exercise 2: Geospatial Data Wrangling"
date: "`r Sys.Date()`"  
format: html
execute: 
  message: false
  warning: false
editor: visual
---

## Setting the Scene

Water is an important resource to mankind. Clean and accessible water is critical to human health. It provides a healthy environment, a sustainable economy, reduces poverty and ensures peace and security. Yet over 40% of the global population does not have access to sufficient clean water. By 2025, 1.8 billion people will be living in countries or regions with absolute water scarcity, according to UN-Water. The lack of water poses a major threat to several sectors, including food security. Agriculture uses about 70% of the world's accessible freshwater.

Developing countries are most affected by water shortages and poor water quality. Up to 80% of illnesses in the developing world are linked to inadequate water and sanitation. Despite technological advancement, providing clean water to the rural community is still a major development issues in many countries globally, especially countries in the Africa continent.

To address the issue of providing clean and sustainable water supply to the rural community, a global [Water Point Data Exchange (WPdx)](https://www.waterpointdata.org/about/) project has been initiated. The main aim of this initiative is to collect water point related data from rural areas at the water point or small water scheme level and share the data via WPdx Data Repository, a cloud-based data library. What is so special of this project is that data are collected based on [WPDx Data Standard](https://www.waterpointdata.org/wp-content/uploads/2021/04/WPDx_Data_Standard.pdf).

## Objectives

Geospatial analytics hold tremendous potential to address complex problems facing society. In this study, you are tasked to apply appropriate geospatial data wrangling methods to prepare the data for water point mapping study. For the purpose of this study, Nigeria will be used as the study country.

## The Data

### Apstial data

For the purpose of this assignment, data from [WPdx Global Data Repositories](https://www.waterpointdata.org/access-data/) will be used. There are two versions of the data. They are: WPdx-Basic and WPdx+. You are required to use WPdx+ data set.

### Geospatial data

Nigeria Level-2 Administrative Boundary (also known as Local Government Area) polygon features GIS data will be used in this take-home exercise. The data can be downloaded either from The [Humanitarian Data Exchange](https://data.humdata.org/) portal or [geoBoundaries](https://www.geoboundaries.org/index.html).

## The Task

The specific tasks of this take-home exercise are as follows:

-   Using appropriate sf method, import the shapefile into R and save it in a simple feature data frame format. Note that there are three Projected Coordinate Systems of Nigeria, they are: EPSG: 26391, 26392, and 26303. You can use any one of them.
-   Using appropriate tidyr and dplyr methods, derive the proportion of functional and non-functional water point at LGA level.
-   Combining the geospatial and aspatial data frame into simple feature data frame.
-   Visualising the distribution of water point by using appropriate analytical visualisation methods.

## Getting started

For the purpose of this in-class exercise, three R packages will be used. They are: sf, tidyverse and funModeling.

::: callout-note
## Your turn

Using the step you had learned, check if these three R packages have been installed in you laptop, if not install the missing R packages. If Yes, launch the R packages into R environment
:::

```{r}
#| code-fold: true
#| code-summary: "Show the code"
pacman::p_load(sf, tidyverse, funModeling)
```

## Importing Geospatial

::: callout-note
## Your turn

Using the step you had learned, import the LGA boundary GIS data of Nigeria downloaded from both sources recommend above.
:::

### The geoBoundaries data set

```{r}
#| code-fold: true
#| code-summary: "Show the code"
geoNGA <- st_read("data/geospatial/",
                  layer = "geoBoundaries-NGA-ADM2") %>%
  st_transform(crs = 26392)
```

### The NGA data set

```{r}
#| code-fold: true
#| code-summary: "Show the code"
NGA <- st_read("data/geospatial/",
               layer = "nga_admbnda_adm2") %>%
  st_transform(crs = 26392)
```

## Importing Aspatial data

::: callout-note
### Your turn

Using the steps you had learned, import the downloaded water point data set into R.
:::

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| eval: false
wp_nga <- read_csv("data/aspatial/WPdx.csv") %>%
  filter(`#clean_country_name` == "Nigeria")
```

### Write the extracted data into rds format

```{r}
#| code-fold: true
#| code-summary: "Show the code"
```

## Converting Aspatial Data into Geospatial

::: callout-note
### Your turn

Using the steps you had learned, convert the newly extracted wp_NGA into point sf data frame
:::

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| eval: false
wp_nga$Geometry = st_as_sfc(wp_nga$`New Georeferenced Column`)
wp_nga
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| eval: false
wp_sf <- st_sf(wp_nga, crs=4326)
wp_sf
```

### Projection transformation

::: callout-note
### Your turn

Using the steps you had learned, transform the projection from wgs84 to appropriate projected coordinate system of Nigeria.
:::

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| eval: false
wp_sf <- wp_sf %>%
  st_transform(crs = 26392)
```
