---
title: "In-class Exercise 10: Spatial Interaction Models with R"
author: "Dr. Kam Tin Seong"
---

```{r}
pacman::p_load(sf, accessibility, tidyverse)
```

```{r}
data_dir <- system.file("extdata", package = "accessibility")

travel_matrix <- readRDS(file.path(data_dir, "travel_matrix.rds"))
head(travel_matrix)
```

The land use data must also be structured in a data.frame and must contain an id column, referring to the ids listed in the travel matrix, and the number of opportunities/facilities/services in each spatial unit. The sample dataset we’ll be using looks like this:

```{r}
land_use_data <- readRDS(file.path(data_dir, "land_use_data.rds"))
head(land_use_data)
```

Gravity measures

gravity() calculates gravity-based measures - i.e. measures in which the weight of opportunities is gradually discounted as the travel cost increases. Of course, several different decay functions can be used to so, each one of them with a range of possible different parameters. In order to accommodate such generalization, the function takes the decay function to be used as a parameter.

In the example below, we calculate accessibility using a negative exponential function with a decay_value (usually referred as the in its formulation) of 0.2. Please see the vignette on decay functions for more information on the decay functions shipped with the package and how to use custom functions.

```{r}
negative_exp <- gravity(
  travel_matrix,
  land_use_data,
  opportunity = "schools",
  travel_cost = "travel_time",
  decay_function = decay_exponential(decay_value = 0.2)
)
head(negative_exp)
```